SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-09-29 response/response.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{dcj}

\begin{document}

We appreciate the time and effort extended by the anonymous reviewers and the
editorial staff to evaluate the original version of this manuscript, which was
submitted on July 8. Upon review, the second of the two reviewers raised some
concerns leading to its rejection. However, we believe that the bulk of these
concerns are either incongruous with the content of the manuscript or
addressable with very little effort, and therefore do not warrant its rejection.
Here we have provided a detailed response to these comments, and have included
additional material in an effort to clarify these points.

The second reviewer raised three major concerns. We summarize our
responses to these issues here.  The first concern raised is
that the method we present is unjustifiably complex, involving a ``plethora of
tuning parameters'', and that the much simpler method of Hansen, et al., would
perform as well if trained on enough data.  In the data we presented, the simple
method developed by Hansen, et al., performed poorly regardless of the number of
reads used in training, which varied between hundreds of thousands to tens of
millions.  The methods presented by Li, et al., are more effective, yet involve
more parameters, and require more time to train, so that our method, while often
being more effective, also offers a significant simplification.
Additionally, we believe the structure of our models give some
theoretical insight into these differences in efficacy.

Secondly, it is asserted that we did not demonstrate that uniformity of read
coverage is increased after correcting for bias using our method.  Yet, in
section 3.2 we had presented a rigorous statistical analysis over a large and
diverse collection of data sets showing that this is indeed the case. This
section of the paper was perhaps overlooked by the reviewer, as it received
no mention in this criticism.

@@

Finally, the reviewer doubts the biological relevance of the bias correction
methods. Having demonstrated an increase in uniformity (section 3.2) and a
decrease in nucleotide frequency divergence (section 3.1), we believe a
compelling case was made that correcting for bias increases the accuracy of
quantification. In this revision we have supplied another line of evidence:
analysis showing that correlation between RNA-Seq and qRT-PCR is increased after
correcting for bias.

Under these circumstances we respectively request that the manuscript, in this
revised form, be reconsidered for publication.  Detailed responses to
the reviewers concerns are presented below.

@@

\subsection*{Response to Reviewer 1}

\begin{quote}
(1) Page 3, 2.2.1 mentions a correction for bias introduced by PCR amplification.
RNA sequencing methods are available where PCR amplification is not used.
Correcting for bias in these cases removes true duplication.  Could the authors
please comment on this?
\end{quote}

First, we emphasize that PCR bias is only one of a number of technical steps
involved in sequencing that potentially introduce bias.  Our method is agnostic
to the source of bias, and because model complexity is penalized, a very sparse,
or even empty model will be trained on a data set without bias.  Thus, if PCR
amplification is responsible for a majority of the bias, and an RNA-Seq method
without PCR amplification is used producing largely bias-free data, then using
the method to correct for bias will have little if any effect.

As a concrete example, we have included new data examining the FRT-Seq protocol
of Mamanova, et al., in which no PCR-amplification is performed during library
preparation. We observe that bias is still present in this data set, albeit at
lower levels than other uncorrected RNAseq data examined.  Our method accounts
for this reduced bias by constructing a much sparser model, relative to the
other RNA-Seq data sets we have examined.

To further explore the point raised, we repeatedly trained our model on sets of
10,000 random simulated reads, drawn uniformly from within exons. We repeated
this 100 times, and in each trial an empty model was trained.  Reducing the
number of reads increases the risk of a spuriously non-empty model, but
repeating this evaluating with only 100 reads led to only 3 out of 100 models
being non-empty. This is indicative that the default BIC scoring, as predicted
by statistical theory, is adequately protecting against false discovery of
``bias'' when there is none.

@@

\subsection*{Response to Reviewer 2}

\begin{quote}
The ms presents an approach to estimating the effect of local sequence
context on the rate of seeing aligned reads at each position within a
transcript in RNA-Seq data. The method is inspired by Hansen et al.'s,
which simply determines this bias by comparing heptamer frequencies.

Compared to the brute force approach of Hansen, who estimated model
parameters (the density of a joint distribution) directly from
heptamer frequencies, here a Bayesian network approach is taken, which
allows for a sparser parameterisation. However, the cost for this are
a plethora of tuning parameters (e.g. the complexity penalty, data
sampling, $p_{max}$, $d_{max}$) and a greedy algorithm with apparently no
control of the quality of its solution.
\end{quote}

The method has precisely the following six parameters,
\begin{enumerate}
\item The number of reads the method is trained on.
\item The size of the window surrounding read starts over which the model should
be trained.
\item The maximum in-degree (i.e., $p_{\text{max}}$)
\item The maximum edge distance (i.e., $d_{\text{max}}$).
\item The standard deviation at which background sequences are sampled.
\item The model complexity penalty.
\end{enumerate}

Parameters (1), (2), (3), and (4) are introduced only to reduce the CPU time
needed to train the algorithm. Each has the same very simple interpretation: a
larger value leads to a potentially better solution, at the cost of increased
training time.

We mention these parameters in the manuscript only for the sake of transparency.
They generally require no intervention or tuning, being essentially an
implementation detail.

Parameters (5) and (6) are more important to the quality of the solution, yet we
have demonstrated, with no special tuning, the method is effective across a
diverse collection of data sets. In this revision we have included new data
(Supplementary Section 2) showing explicitly that the quality of the solution is
remarkably insensitive to values of these two parameters over a range spanning 4
orders of magnitude.

It should also be noted that, while the brute-force method of Hansen, et. al.,
has fewer tuning parameters, the other methods evaluated in the manuscript (the
GLM and MART methods of Li, et. al.), involve more parameters than the method we
have described. (For example, training the MART method involves parameters
controlling the complexity penalty, the number of trees, the lower bound
threshold for read counts, in addition to others that are not described or
easily changed from the values divined by the authors.  Additionally, their
method requires selection of a set of annotated genes for training.)  Our method
therefore presents a simplification over competing methods.

@@

\begin{quote}
To justify this cost, the
authors make the argument that their approach avoids the potential
overfitting of a brute force approach such as Hansen's. However, I am
not convinced: they evaluate both methods, theirs and Hansen's, on
'more than 600,000 reads', yet this seems unrealistic: current
datasets have hundreds of millions of reads, and that may just solve
the estimation precision problem in the simple brute force approach.

(I appreciate that the implementation of the brute force approach in
the Genominator package is ridiculously inefficient, but that can
easily be fixed by more competent programming.)
\end{quote}

We regret that we were not more explicit on this point. At least
600,000 reads on each data set, but as many as 23 million were used:
see the revised table X. (As noted in the paper, we did reimplement
the Genominator method in order to do this.)

We do not wish to ``pick on'' Hansen, et al.  Theirs was a pioneering
paper and presents a simple, intuitive approach for tackling the
problem they identified.  Nevertheless, their brute-force method
performed poorly on every data set we tested, even the largest ones,
suggesting that simply throwing more data at a suboptimal solution
will not adequately solve the problem.  In our opinion, the results we
presented were unambiguous: the other methods perform far better with
far fewer reads.

%Moreover, even if the brute-force method were effective if trained
%with many hundreds of millions of reads, many (possibly most) existing
%data sets are not this large, rendering it generally impractical.

Understanding this failure of a simple, appealing method is important
for making progress, and we feel that our paper provides a theoretical
basis for understanding these differences in efficacy, and we have
amplified these points in the revision.  Intuitively, one expects
model accuracy to increase as the number $n$ of training reads
increases, but to decrease as the number $r$ of model parameters
grows, since the training data is ``spread more thinly.''  For the
particular case of estimating a multinomial distribution (as in
Hansen), Supplementary Section YY quantifies this intuition very
precisely, showing that the expected KL divergence of the learned
model with respect to the true distribution converges to $(r-1)/(2n)$,
where $r=4^7=16384$ is the number of model parameters learned.  The
same analysis does not directly apply to the MART, GLM or BN models,
but we expect their behavior to be qualitatively similar.  Thus, to
the extent that a more parsimonious model such as ours adequately
captures the structure of the data, it is potentially orders of
magnitude more accurate for a given quantity of training data, simply
by virtue of having only a few hundred parameters learned from counts
(Figure 3), as opposed to tens of thousands.  [KL divergence between
two distributions is inversely proportional to the number of samples
one needs to confidently test the hypothesis that the samples arise
from one distribution versus the other; see Supplementary Section XX.]
%% TODO: why the brackets here?

Of course, training data is plentiful in NGS data sets, so,
pragmatically, the above analysis probably identifies a contributor
but not the root cause of Hansen et al.'s relatively poor performance
in our experiments.  Rather, we believe the problem is that bias is
not confined to the initial heptamer.  The model structures
illustrated in Figure 3 show that nucleotides well outside that region
are often biased.  (The randomization study sketched above shows that
our model building algorithm is not prone to spurious discovery of
overly-complex models.)  Furthermore, these examples show that the
read sampling is often biased by positions \emph{in front of the
  read}, which are not currently considered by Hansen.  Their method
could be modified easily to consider $k$-mers spanning the read start,
but the entropy analysis shows that extending it to $k \gg 7$ is
infeasible, since the exponential growth in $r=4^k$ will destroy the
model's accuracy even in large data sets.  E.g., $k \ge 20$ would be
necessary to cover the biased regions in all data sets shown in Figure
3, necessitating trillions of training reads to build an adequate
model by Hansen's method.

@@

\begin{quote}
The major weakness of the ms, in my view, is the choice of 'figures of
merit' that the authors made (Fig.1, Sections 3.1, 3.2). What I would
have expected to see are

1.) Plots closer to the data such as Supplementary Fig.5 of
reference [1]. That figure shows highly uneven coverage of reads
within exons.
- Does this unevenness go away with the proposed adjustment?
\end{quote}

During preparation we chose not to include a figure such as the one suggested
by the reviewer as we believed it would not be legitimately informative. It
would be far too easy to cherry-pick a miniscule genomic interval that makes
even a very shoddy method appear favorable.

Instead, we presented data (in Section 3.2) that answers the
reviewer's question (i.e., ``does this unevenness go away with the
proposed adjustment?'') in rigorous and comprehensive manner. We
implemented a Poisson regression metric, which quantitatively measures
``uniformity'' by calculating the likelihood that the reads sampled
across a given exon were sampled from a uniform distribution.  Using
this, we performed a large cross-validation study proving the
effectiveness of our method at increasing uniformity across thousands
of exons drawn from a diverse group of data sets, rather than
qualitatively across one exon.  In broad terms, our method increases
likelihood of the uniform model by XXX-fold across thousands of exons
(Figure 6).

In this revision, as suggested, we have revised Figure 2 to show an
example of this increase in uniformity.  We believe the figure is a
nice, graphic illustration of the benefits of our method and thank the
reviewer for causing us to reconsider it.  We stress, however, that
this figure represents a minute subset of the analysis that had
previously been presented, and is included only as an illustration.
The results in sections 3.1 and especially in 3.2 are far more
comprehensive.

\begin{quote}
- Also, as is evident from that figure, for some of the experimental
protocols, some regions within highly expressed exons get vanishingly
small read coverage.  These are the really hard situations, and much
more worrying than instances where coverage just fluctuates a bit.
How does the proposed scaling adjustment of Section 2.1 work in these
places (i.e. when $s_{i}$ is very small)?
\end{quote}

%% TODO

%The revised Figure 2 also graphically illustrates the range of
%adjustment factors, which genome-wide are very cool and we hope we can say something soothing about.......

@@

\begin{quote}
Minor points:
------------
1. The ms seems confused between classification and regression.
Reference [2] might be helpful to resolve some of that.
\end{quote}

The term ``classification'' was used in the context of noting that the algorithm
is adapted from one typically used in classification. It is correct to say that
the method here performs regression, and we have tried to clarify this language
in this revision.

\begin{quote}
2. The chosen Bayesian network is rather complicated, and a spectrum
of alternative methods exists between that and the brute force
approach of Hansen (say, lasso or svm regression). In absence of a
more meaningful benchmark, the choice of method seems arbitrary.
\end{quote}

Our motivation for adopting a Bayesian network model is the very
natural probabilistic interpretation that we have described in Section
2.1. We disagree with the reviewer's opinion that Bayesian networks
are inherently more complicated than Lasso regression or support
vector machines.  In this application, the latter two methods would
generate output that is arguably less transparent and interpretable,
and would neither be significantly simpler to implement, nor involve
fewer parameters.  Our hypotheses (above) concerning the reasons for
the weakness of Hansen's method illustrate the benefit of this
approach; we do not believe svm regression, say, would have yielded
similar insights.

@@

Naturally, it is possible that another model might perform as well or better
than the one presented here, yet it is unrealistic or impossible to prove that
this is not case. What we have demonstrated is a model that is well reasoned,
efficiently implemented, and effective at reducing sequence bias in RNA-Seq.

%\subsection*{Other Changes}

%We have modified the title to more clearly acknowledge the existence of
%the prior work upon which we build.

%...more...?

\subsection*{Conclusion}
In summary, we believe the revised manuscript addresses all concerns
raised by the reviewers, and provides a substantive improvement to the
state of the art for an important problem.  We respectfully request
that the manuscript, in this revised form, be reconsidered for
publication.

### 2011-09-29 seqbias-paper.tex
more revisions

Along with $p_{\text{max}}$ and $d_{\text{max}}$, the time required to train the
model is controlled by the number of reads used (i.e., $n$) and the number of
positions surrounding the read start to consider. Increasing any of these
parameters results in a solution that is no worse, and potentially better, at
the expense of increased training time.

The quality of the solution depends also on two other parameters: the standard
deviation at which background sequences are sampled, and the weight applied to
the penalty term of the BIC, yet it is not particularly sensitive to their values
(see, Supplementary Section 2). In all our evaluation, no special tuning of the
parameters was performed, suggesting it can be used effectively across data sets
without any intervention.

@@

\citet{Li2010} recommends that their MART and GLM models be trained using the
100 most abundant genes. We used 1000 exons from chromosomes 1--8, otherwise
chosen in a manner identical to that which was used to select the test exons.
Both the GLM and MART models were trained using the default parameters.

@@

Our own method was trained on the 100,000 reads from chromosomes 1--8, with the
highest number of duplicate reads.

All data sets were mapped using Bowtie \citep{Langmead2009} against,
respectively, the hg19, mm9, rheMac2, and panTro2 genome assemblies obtained
from the UCSC Genome Browser \citep{Karolchik2008}.

@@

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as
$$D_k( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
where the sum is over all $k$-mers.

@@

We have demonstrated that sequence bias can confound, sometimes severely,
quantification in RNA-Seq experiments, and we have introduced an effective
method to account for this bias without the need of existing gene annotations.
The analysis provided demonstrates that our method shows significant improvement
in two aspects: uniformity of read coverage, and consistency of nucleotide
frequencies. These results strongly suggest an increase in accuracy, and we show
in Supplementary Section 4, that this conclusion is further justified by
comparison to qRT-PCR.

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate
heptamer frequencies, a likely explanation is that the model is overfit to the
training set.  We trained the 7mer model on a minimum of 600,000 reads (with the
Bullard data set), and up to 23 million reads (with the Wetterbom data set), yet
it seems more reads, or a greater diversity of reads are needed to produce an
accurate or useful estimate.

Our method generalizes this approach, attempting to overcome this problem by
using an estimation of sequence probability that is more robust to overfitting
and can account for bias outside of the initial heptamer. In all our tests, this
approach was at least as effective as those of \citet{Li2010}, despite not
requiring gene annotations or manual selection of training examples.

>>>>

Along with $p_{\text{max}}$ and $d_{\text{max}}$, the time required to train the
model is controlled by the number of reads used (i.e., $n$) and the number of
positions surrounding the read start to consider. Increasing any of these
parameters results in a  solution that is potentially better, at
the expense of increased training time.

The quality of the solution depends also on two other parameters: the standard
deviation at which background sequences are sampled, and the weight applied to
the penalty term of the BIC, yet it is not particularly sensitive to their
values. (The median $R^2$ goodness-of-fit statistic used in section 3.2 varied
by less that 25\% as these parameters were varied over a range of $10^4$. See
Supplementary Section 2.) In all our evaluation, no special tuning of the
parameters was performed, suggesting it can be used effectively across data sets
without any intervention.

@@

\citet{Li2010} recommends that their MART and GLM models be trained using the
100 most abundant genes. We used 1000 exons from chromosomes 1--8, otherwise
chosen in a manner identical to that which was used to select the test exons.
Both the GLM and MART models were trained considering the initial read position
and 20 nucleotides upstream and downsream, and otherwise using default
parameters.

@@

Our own method was trained on the 100,000 randomly selected reads from
chromosomes 1--8, considering the initial read position and 20 nucleotides
upstream and downstream.

All data sets were mapped using Bowtie \citep{Langmead2009} using default
parameters against, respectively, the hg19, mm9, rheMac2, and panTro2 genome
assemblies obtained from the UCSC Genome Browser \citep{Karolchik2008}.

@@

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as
$$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
where the sum is over all $k$-mers.

@@

We have demonstrated that sequence bias can confound, sometimes severely,
quantification in RNA-Seq experiments, and we have introduced an effective
method to account for this bias without the need of existing gene annotations.
The analysis provided demonstrates that our method shows significant improvement
in three aspects: uniformity of read coverage, and consistency of nucleotide
frequencies, and agreement with qRT-PCR.

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. A possible explaination is, given the large number of
parameters needed to estimate heptamer frequencies ($4^{7} = 16,383$), these
parameters are estimated with less accuracy than in models requiring fewer
parameters. Yet, we trained the 7mer model on a minimum of 600,000 reads (with
the Bullard data set), and up to 23 million reads (with the Wetterbom data set),
numbers which our theoretical results, detailed in Supplementary Section 9,
suggest should lead to accurate estimates. A perhaps more significant factor is
that this method does not capture bias outside of the initial heptamer, though
many data sets clearly are affected by bias in other positions.  However, if we
were to incease the size of the $k$-mer being considered, exponentially more
reads would be required for an accurate estimate.

Our method generalizes this approach, attempting to overcome this problem by
using an estimation of sequence probability that requires fewer parameters
and can account for bias outside of the initial heptamer. In all our tests, this
approach was at least as effective as those of \citet{Li2010}, despite not
requiring gene annotations or manual selection of training examples.

### 2011-09-29 seqbias-supplement.tex
more revisions

\documentclass[letterpaper]{article}
\usepackage{dcj}
\usepackage{lscape}
\usepackage{comment}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{fontspec}

@@

Observing the nonuniform distribution of nucleotide frequencies surrounding the
5' end of reads, a natural step to take would be to trim the 5' end before mapping,
in the hope that simply removing the portion of the read in which the bias
occurs will also remove the bias. The figure below demonstrates that this is not
the case. Trimming the initial heptamer in the Mortazavi data set does nothing to
reduce the bias, and simply shifts the plot by seven positions. This indicates
that the issue is \emph{sampling bias}, rather than a bias in base calling.

\begin{center}
\includegraphics[width=0.35\textwidth]{fig/trimmed-freqs.pdf}
\end{center}

@@

\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-sampstd.pdf}
\end{center}

The above plot show the standard deviation used to draw background samples
varied from 1 to 10,000. Apparent from this plot is that, while the optimal
choice lies somewhere between 10 and 100, the model performs competitively for
any reasonable value.

\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-comppen.pdf}
\end{center}

Here we multiply the complexity penalty term of the BIC by constants varying
from $10^{-6}$ to $10^{3}$. (Recall that the BIC is $2\ell - m \log n$ where
$\ell$ is the log-likelihood, $m$ is the number of parameters needed to specify
the model, and $n$ is the number of training examples. Here we compute $2 \ell -
c m \log n$, varying $c$.)

It is clear from the above figure, that a very severe complexity penalty will
result in a poor model. Here, for constants larger than 100, an empty model is
trained, resulting in the goodness of fit of 0. Conversely, if this constant is
very small, a overly-dense model will be trained.  In this data, for constants
less than 0.01, the maximally dense model is trained (given the restraints on
in-degree and edge distance). This result is a sub-optimal but entirely adequate
solution.

@@

Below we plot the adjusted and unadjusted KL divergence plots, supplementing
those in the results section of the main paper.

\includegraphics[width=\textwidth]{fig/kl-all.pdf}

@@

\centerline{\includegraphics[width=0.6\textwidth]{fig/au-freqs.pdf}}

@@

Training our model on more reads results in more accurate estimation of bias,
but a longer training time. Here we investigate that trade-off by training our
model on progressively more reads from the Mortazavi data set. We use the
median pseudo-coefficient of determination ($R^2$) over exons selected for our
test set. Each point is additionally labeled with the time used to train the
model on one core of a 3Ghz Intel Xeon processor.

\includegraphics[width=\textwidth]{fig/pois-n.eps}

@@

\begin{center}
\includegraphics[width=0.7\textwidth]{fig/freqs_cao.pdf}
\end{center}

@@

\begin{center}
\includegraphics[width=0.5\textwidth]{fig/cao-kl.pdf}
\end{center}

@@

\begin{center}
\includegraphics[width=0.6\textwidth]{fig/frt-all.pdf}
\end{center}

@@

\begin{center}
\includegraphics[width=0.4\textwidth]{fig/frt-graph.pdf}
\end{center}

@@

\begin{center}
\includegraphics[width=0.5\textwidth]{fig/replicates.pdf}
\end{center}

@@

The summation is the test statistic for the $\chi^2$ goodness-of-fit test for a
multinomial distribution, and as $n \rightarrow \infty$ is known to follow a
$\chi^2$ distribution with $r - 1$ degrees of freedom. Finally, the expected
value of such a random variable is $r - 1$, hence the expected KL divergence of
the MLE inferred distribution $Q$ with respect to the true distribution $P$ is
$$E[H(Q||P)] = \frac{r - 1}{2n}$$

Stochastic simulations of this proves when $P$ is uniform show that the above
formula is a very good fit when $n > r$.
%% TODO: see figure X

>>>>

\documentclass[letterpaper]{article}
\usepackage{dcj}
\usepackage{lscape}
\usepackage{comment}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{fontspec}
\usepackage{float}

@@

Observing the nonuniform distribution of nucleotide frequencies surrounding the
5' end of reads, a natural step to take would be to trim the 5' end before
mapping, in the hope that simply removing the portion of the read in which the
bias occurs will also remove the bias. Figure \ref{fig:trimmedfreqs}
demonstrates that this is not the case. Trimming the initial heptamer in the
Mortazavi data set does nothing to reduce the bias, and simply shifts the plot
by seven positions. This indicates that the issue is \emph{sampling bias},
rather than a bias in base calling.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.35\textwidth]{fig/trimmed-freqs.pdf}
\end{center}
\caption{Nucleotide frequencies observed before (plotted in red) and after
(plotted in black) trimming the initial heptamer.}
\label{fig:trimmedfreqs}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-sampstd.pdf}
\end{center}
\caption{Sensitivity of the parameter control he standard deviation at which
background sequences are sampled.}
\label{fig:poissampstd}
\end{figure}

Figure \ref{fig:poissampstd} show the standard deviation used to draw background
samples varied from 1 to 10,000. Apparent from this plot is that, while the
optimal choice lies somewhere between 10 and 100, the model performs
competitively for any reasonable value.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-comppen.pdf}
\end{center}
\caption{Sensitivity over adjustments to the complexity penalty.}
\label{fig:poiscomppen}
\end{figure}

Next we multiplied the complexity penalty term of the BIC by a constants varying
from $10^{-6}$ to $10^{3}$. (Recall that the BIC is $2\ell - m \log n$ where
$\ell$ is the log-likelihood, $m$ is the number of parameters needed to specify
the model, and $n$ is the number of training examples. Here we compute $2 \ell -
c m \log n$, varying $c$.)

It is clear from Figure \ref{fig:poiscomppen}, that a very severe complexity
penalty will result in a poor model. Here, for constants larger than 100, an
empty model is trained, resulting in a median $R^2$ of 0. Conversely, if
this constant is very small, a overly-dense model will be trained.  In this
data, for constants less than 0.01, the maximally dense model is trained (given
the restraints on in-degree and edge distance). This result is a sub-optimal but
entirely adequate solution.

@@

Figure \ref{fig:klall} plots the adjusted and unadjusted KL divergence plots,
supplementing those in the results section of the main paper.

\begin{figure}[H]
\centerline{
\includegraphics[width=\textwidth]{fig/kl-all.pdf}
}
\caption{}
\label{fig:klall}
\end{figure}

@@

Figure \ref{fig:aufreqs} plots sequence bias and KL divergence in the data
published by Au, et al. \cite{Au2010} and examined by us in Section 3.3.

\begin{figure}[H]
\centerline{\includegraphics[width=0.6\textwidth]{fig/au-freqs.pdf}}
\caption{}
\label{fig:aufreqs}
\end{figure}

@@

Training our model on more reads results in more accurate estimation of bias,
but a longer training time. Here we investigate that trade-off by training our
model on progressively more reads from the Mortazavi data set. For each subset
we train our method, evaluate the median pseudo-coefficient of determination
($R^2$) over exons selected for our test set, and record the training time
required on one core of a 3Ghz Intel Xeon processor. These results are plotted
in Figure \ref{fig:poisn}.

\begin{figure}[H]
\includegraphics[width=\textwidth]{fig/pois-n.eps}
\caption{Median $R^2$ is plotted against training set size. Each point is
additionally labeled with the run time of the training procedure.}
\label{fig:poisn}
\end{figure}

%% TODO: coptions/labels below this point

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig/freqs_cao.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.5\textwidth]{fig/cao-kl.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.6\textwidth]{fig/frt-all.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.4\textwidth]{fig/frt-graph.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig/replicates.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

The summation is the test statistic for the $\chi^2$ goodness-of-fit test for a
multinomial distribution, and as $n \rightarrow \infty$ is known to follow a
$\chi^2$ distribution with $r - 1$ degrees of freedom. Finally, the expected
value of such a random variable is $r - 1$, hence the expected KL divergence of
the MLE inferred distribution $Q$ with respect to the true distribution $P$ is
\begin{equation}
\label{eq:entropy}
E[H(Q||P)] = \frac{r - 1}{2n}
\end{equation}

Stochastic simulations of this proves when $P$ is uniform show that the above
formula is a very good fit when $n > r$.

\begin{figure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{fig/entropy.pdf}}
\caption{For each value of $n$ plotted, the circles indicate the value of
$H(Q||P)$, averaged over 100 samples of size $n$, where $P$ is uniform and $Q$
is the MLE estimator for $P$ basen on $n$ random samples drawn from $r$ bins.
The asterisks and the straight line interpolating them are the theoretical
approximation from Equation \ref{eq:entropy}.
}
\label{fig:entropy}
\end{figure}
